# Welcome

This guide combines the learning for both *AS 1.1* (91944: Explore data using a statistical enquiry process) and *AS 1.3* (91946: Interpret and apply statistical information). By highlighting where these two standards overlap, this resource eliminates unnecessary duplication and shows you how core statistical thinking connects the data you collect with the data you evaluate in everyday life.

## One toolkit, two assessments

You are not learning two separate subjects. The same toolkit appears in both
courses:

- **AS1.1 (Internal)** - statistical investigation in context
- **AS1.3 (External / 91946)** - interpret and apply statistical information in context

In the internal assessment, you use that toolkit to plan an investigation,
collect data, analyse it, and write a conclusion. In the external assessment,
you use the same ideas to read graphs, interpret tables, question claims, and
justify your reasoning under exam conditions.

## Where the standards overlap

| Area | AS1.1 Internal | AS1.3 External (91946) | Overlap? |
|------|----------------|------------------------|----------|
| Statistical enquiry cycle (PPDAC) | Core assessed process | Used for interpretation and critique | **Both** |
| Designing and carrying out an investigation | Strong focus | Not the primary assessment focus | **Mostly 1.1** |
| Reading graphs/tables in context | Required | Required | **Both** |
| Numerical summaries (mean, median, IQR, spread) | Used for evidence in investigation | Directly assessed in interpretation questions | **Both** |
| Probability, two-way tables, expected value | Limited or indirect | Directly assessed | **Mostly 1.3** |
| Justification and statistical reasoning | Required in conclusion and evaluation | Required in written responses | **Both** |
| Exam-style interpretation under time pressure | Not the main format | Core format | **Mostly 1.3** |

## How this book is organised

This draft keeps the shared statistical foundation together first, then moves
probability and critical reading into a second part.

| Part | Main focus | Why it comes here |
|------|------------|-------------------|
| **Part 1** | Statistical questions, data types, displays, comparison, and practice | These chapters support both standards immediately. |
| **Part 2** | Probability and critical reading of statistical information | These ideas matter for AS1.3, but they can sit after the core foundation. |

## Assessment tags used in this book

- **[1.1]** mainly assessed in the internal
- **[1.3]** mainly assessed in the external
- **[Both]** directly useful for both standards

## Suggested learning pathway

Follow this sequence from start to finish:

| Order | Learning focus | Where in this book | Tested in |
|------:|----------------|--------------------|-----------|
| 1 | Combined overview of AS1.1 and AS1.3 expectations | Chapter 1 - Combined approach | **[Both]** |
| 2 | Statistical questions, context language, core terms | Chapter 2 - What is Statistics? | **[Both]** |
| 3 | Statistical enquiry cycle and investigation flow | Chapter 3 - PPDAC | **[Both]** |
| 4 | Variables and data types, choosing suitable displays | Chapter 4 - Data types | **[Both]** |
| 5 | Independent and dependent variable roles in investigations | Chapter 5 - Independent and Dependent Variables | **[Both]** |
| 6 | Spread-focused notes with quartiles and IQR worked examples | Chapter 6 - Measures of Spread Study Notes | **[Both]** |
| 7 | Graph interpretation and comparison language | Chapter 7 - Graphs | **[Both]** |
| 8 | Applied investigation example in context | Chapter 8 - Koura Example | **[Both]** |
| 9 | Centre and spread for comparison and conclusions | Chapter 9 - Measures | **[Both]** |
| 10 | Applied tasks and mixed practice with response scaffolds | Chapter 10 - Practice | **[Both]** |
| 11 | Probability tools and table-based reasoning | Part 2 - Probability | **[1.3]** |
| 12 | Sampling quality, bias, inference, critique | Part 2 - Critical reading | **[Both]** |
| 13 | Extension reading and source material | Resource Library | **[Both]** |

## How to use this pathway for dual preparation

1. Build investigation capability first (Chapters 1-6) for strong **AS1.1** evidence.
2. Add variable-role clarity with the independent/dependent chapter, then
	consolidate spread ideas in the measures-of-spread notes chapter.
3. Use the graphs, koura example, measures, and practice chapters to
	consolidate comparison and communication.
4. Add the Part 2 extension chapters for explicit **AS1.3** preparation.
5. In every written response, include context, comparison, and justification language.

## What this standard is about

In AS1.3 (91946), you will learn to:

- **Read and interpret** statistical displays, tables, and graphs
- **Apply** mathematical and statistical ideas to real-world problems
- **Communicate** findings clearly using statistical language
- **Evaluate** statistical claims made in media and everyday life

## How to use this book

Each chapter builds on the previous one. You will find:

- Clear explanations of key concepts with New Zealand examples
- Interactive R code blocks in the HTML version showing how statistics is done with real data
- Practice questions at Achievement, Merit, and Excellence levels
- Glossary boxes highlighting important vocabulary

In the PDF version, the working code is hidden so the focus stays on the
mathematics and interpretation. In the HTML version, the code can still be
opened for teaching, demonstration, or extension.

::: {.callout-tip}
## At a Glance: AS1.3 (External)

| Grade | What you need to do |
|-------|---------------------|
| **Achieved** | Interpret and apply statistical information |
| **Merit** | Interpret and apply statistical information with justification |
| **Excellence** | Interpret and apply statistical information with statistical insight |
:::

::: {.callout-note}
## Internal/External connection
If you can explain findings from your own investigation clearly and in context
(AS1.1), you are building the same reasoning needed for interpretation and
justification in AS1.3.
:::

## Tools we use

All data visualisations and calculations in this book are produced using **R** and
rendered with **Quarto**. You do not need to be able to write R code to succeed
in this standard. The code is there so your teacher can demonstrate the
analysis live, and so students who want extension material can inspect how the
graphs and summaries were built.

```{r}
#| label: welcome-setup
#| include: false

source("book-support.R")
```

```{r}
#| label: welcome-plot
#| fig-cap: "An example of the kind of visualisations we will create in this book."
#| echo: false

plot_welcome_weather()
```

## If statistics feels new

Start with the core chapters and keep a small set of support resources nearby:

- Khan Academy for quick refreshers on averages, graphs, and probability
- CensusAtSchool New Zealand for local student data examples
- Stats NZ for authentic graphs, tables, and media-style summaries
- The Resource Library at the end of this book for curriculum and standard documents

## Getting started

Head to @sec-intro to begin the combined 1.1 + 1.3 pathway.
