# What is Statistics? {#sec-intro}

```{r}
#| label: intro-setup
#| include: false

source("../book-support.R")
```

```{r}
#| label: fig-statistics-poster
#| fig-cap: "Statistics connects questions, data, and decisions in real contexts."
#| echo: false

plot_statistics_poster()
```

## Why statistics matters

Every day we are surrounded by data. News headlines cite poll results, sports
commentators compare player averages, health campaigns report risk percentages,
and social media shows us trending topics. **Statistics** is the science that
helps us make sense of all this information.

At school, statistics helps you ask better questions, decide what evidence is
relevant, and explain what the data does or does not support. Outside school,
it helps you read media claims critically instead of accepting every graph or
headline at face value.

::: {.callout-note}
## Key definition
**Statistics** is the practice of collecting, organising, summarising,
analysing, and interpreting data in order to draw conclusions and make
decisions.
:::

## Statistical questions and variability

Not all questions are statistical questions. A statistical question expects
**variability** in the answers.

::: {.callout-tip}
**Statistical question:** "How tall are the students in Year 11?"  
*(Heights will vary - this is statistical.)*

**Not a statistical question:** "How tall is Ms Smith?"  
*(Only one answer - not statistical.)*
:::

### Practice identifying statistical questions

Which of the following are statistical questions?

1. How many pages are in Harry Potter and the Philosopher's Stone?
2. How many hours per week do Year 11 students spend on homework?
3. What is the population of New Zealand?
4. What are the most popular music genres among teenagers in Auckland?
5. What colour is the school uniform?
6. How long does it take students in our class to get to school?
7. Which lunch option is chosen most often in the canteen each Friday?
8. How many goals did the first XI football team score this season?
9. What is the average screen time for Year 11 students on a school night?
10. How many desks are in Room 12?

::: {.callout-note .content-visible collapse="true" when-format="html"}
## Answer key
1. Not statistical - there is one fixed answer.
2. Statistical - different students spend different amounts of time.
3. Not statistical - there is one (changing but definite) answer.
4. Statistical - different teenagers have different preferences.
5. Not statistical - the uniform colour is fixed.
6. Statistical - travel times will vary between students.
7. Statistical - different students make different choices.
8. Not statistical - the season total is one fixed value.
9. Statistical - students will report different screen times.
10. Not statistical - there is one fixed count for that room.
:::

### A template for writing your own statistical question

Use this sentence frame when you are stuck:

> What is the distribution of **[variable]** for **[population]** in **[context]**?

For example:

- What is the distribution of weekly study time for Year 11 students at our school?
- How does travel time to school differ for students who take the bus compared with students who walk?

## Population, sample, and variable

These three ideas appear constantly in both the internal and the external.

| Idea | What it means | School-based example |
|------|----------------|----------------------|
| **Population** | The full group you want to know about | All Year 11 students at the school |
| **Sample** | The smaller group you actually collect data from | The 30 Year 11 students surveyed in form time |
| **Variable** | What you measure or record | Height, travel time, or favourite sport |

::: {.callout-warning}
## Common trap
A large sample is not automatically a good sample. If the sample is biased,
the conclusion can still be misleading.
:::

## Why context matters

In AS91946, **context** is central. You will always work with data that
relates to real situations. When interpreting statistics, you must:

- **Name the variable** being measured
- **State the units** (e.g. centimetres, dollars, minutes)
- **Refer to the population** the data came from
- **Relate findings back** to the real-world situation

The strongest answers do more than quote a number. They explain what that
number means for the people, objects, or events being studied.

::: {.callout-warning}
## Common mistake
Saying "the mean is 24" is not enough. Say  
*"The mean age of survey respondents was 24 years."*
:::

### Population vs sample in context

Suppose a school wants to know how much sleep Year 11 students get on a normal
school night.

- The **population** is all Year 11 students at the school.
- A **sample** might be one statistics class or 40 students chosen at random.
- The **variable** is hours of sleep on a school night.

If the sample only comes from one extension class, the results might not
represent the whole year level. That is why context and sampling always matter
when you interpret a summary or graph.

## Glossary

| Term | Meaning |
|------|---------|
| **Data** | Values collected by observation or measurement |
| **Variable** | A characteristic that can take different values |
| **Population** | The entire group of interest |
| **Sample** | A subset of the population selected for study |
| **Statistic** | A numerical summary calculated from a sample |
| **Parameter** | A numerical summary of a whole population |
| **Context** | The real-world situation that gives meaning to the data |

## Exit ticket

1. Give two examples of statistical questions you could investigate at your school.
2. A student says "the average score is 65". What is missing from this statement?
3. What is the difference between a population and a sample?
4. Write one sentence that uses a summary statistic in full context.
5. Explain why a sample could be large but still poor quality.

::: {.callout-note .content-visible collapse="true" when-format="html"}
## Exit ticket answers
1. Any valid examples should expect variation, such as study time, travel time, or sleep.
2. The variable, the units, and the group being described are missing.
3. The population is the whole group of interest; the sample is the smaller group actually measured.
4. Example: "The median travel time for Year 11 students in our survey was 18 minutes."
5. A large sample can still be biased if it leaves out important groups or is collected unfairly.
:::
