# Types of Data and Variables {#sec-data-types}

Understanding the **type of data** you have is the first step in choosing the
right graph or summary statistic. The wrong choice can make data hard to
understand — or even misleading.

## Categorical vs numerical data

```{r}
#| label: data-types-tree
#| fig-cap: "Classification of data types."
#| echo: false

# Simple text-based tree using plot
plot(0, 0, xlim = c(0, 10), ylim = c(0, 6), type = "n",
     axes = FALSE, xlab = "", ylab = "", main = "Types of Data")

# Data node
rect(3.5, 5, 6.5, 5.8, col = "#2c7bb6", border = NA)
text(5, 5.4, "Data", col = "white", font = 2)

# Categorical node
rect(0.5, 3, 3.5, 3.8, col = "#4dac26", border = NA)
text(2, 3.4, "Categorical", col = "white", font = 2)

# Numerical node
rect(6.5, 3, 9.5, 3.8, col = "#d7191c", border = NA)
text(8, 3.4, "Numerical", col = "white", font = 2)

# Nominal
rect(0, 1, 2, 1.8, col = "#b8e186", border = NA)
text(1, 1.4, "Nominal", font = 2, cex = 0.9)

# Ordinal
rect(2.2, 1, 4.2, 1.8, col = "#b8e186", border = NA)
text(3.2, 1.4, "Ordinal", font = 2, cex = 0.9)

# Discrete
rect(5.5, 1, 7.5, 1.8, col = "#f4a582", border = NA)
text(6.5, 1.4, "Discrete", font = 2, cex = 0.9)

# Continuous
rect(7.8, 1, 9.8, 1.8, col = "#f4a582", border = NA)
text(8.8, 1.4, "Continuous", font = 2, cex = 0.9)

# Connecting lines
segments(5, 5, 2, 3.8, lwd = 1.5)
segments(5, 5, 8, 3.8, lwd = 1.5)
segments(2, 3, 1, 1.8, lwd = 1.5)
segments(2, 3, 3.2, 1.8, lwd = 1.5)
segments(8, 3, 6.5, 1.8, lwd = 1.5)
segments(8, 3, 8.8, 1.8, lwd = 1.5)
```

## Categorical data

**Categorical (qualitative) data** describes qualities or categories. The
values are labels, not numbers.

### Nominal data

Nominal data has **no natural order**. Categories are just names.

**Examples:**
- Eye colour (brown, blue, green, grey)
- Favourite sport (rugby, netball, football, basketball)
- Type of transport to school (bus, car, walk, bike)

### Ordinal data

Ordinal data has a **natural order**, but the gaps between categories are not
equal.

**Examples:**
- Survey ratings (strongly disagree → disagree → neutral → agree → strongly agree)
- Year group (Year 9, Year 10, Year 11, Year 12, Year 13)
- Clothing size (XS, S, M, L, XL)

## Numerical data

**Numerical (quantitative) data** represents quantities that can be measured
or counted.

### Discrete data

Discrete data can only take **specific, separate values** (usually whole
numbers from counting).

**Examples:**
- Number of siblings
- Number of text messages sent per day
- Goals scored in a rugby match

### Continuous data

Continuous data can take **any value within a range** (usually from measuring).

**Examples:**
- Height (cm)
- Weight (kg)
- Time to complete a race (seconds)
- Temperature (°C)

## Real data example

```{r}
#| label: data-types-example
#| tbl-cap: "Example student survey data — identify the type of each variable."

# Simulated class survey data
set.seed(7)
n <- 10
survey <- data.frame(
  Student   = paste0("S", 1:n),
  Gender    = sample(c("Male", "Female", "Non-binary"), n, replace = TRUE),
  Height_cm = round(rnorm(n, 168, 10), 1),
  Siblings  = sample(0:4, n, replace = TRUE),
  Transport = sample(c("Bus", "Car", "Walk", "Bike"), n, replace = TRUE),
  Rating    = sample(c("Poor", "Fair", "Good", "Excellent"), n, replace = TRUE)
)

knitr::kable(survey, align = "c")
```

::: {.callout-tip}
## Identify the variable types
- **Gender** → Categorical, Nominal
- **Height_cm** → Numerical, Continuous
- **Siblings** → Numerical, Discrete
- **Transport** → Categorical, Nominal
- **Rating** → Categorical, Ordinal
:::

## Choosing the right display

The type of data determines the best graph to use:

| Data type | Suitable displays |
|-----------|------------------|
| Categorical (nominal/ordinal) | Bar chart, pie chart, dot plot |
| Numerical (discrete) | Dot plot, bar chart, stem-and-leaf |
| Numerical (continuous) | Histogram, box plot, dot plot, stem-and-leaf |
| Two numerical variables | Scatter plot |

## Glossary

| Term | Meaning |
|------|---------|
| **Variable** | A characteristic that varies between individuals |
| **Categorical** | Data that falls into named groups or categories |
| **Nominal** | Categorical with no natural order |
| **Ordinal** | Categorical with a natural order |
| **Numerical** | Data that can be measured or counted |
| **Discrete** | Numerical data that can only take specific values |
| **Continuous** | Numerical data that can take any value in a range |

## Check your understanding

Classify each variable as **categorical (nominal/ordinal)** or **numerical
(discrete/continuous)**:

1. Shoe size
2. Favourite subject at school
3. Time spent sleeping (hours per night)
4. Number of books read last month
5. Blood type
6. Temperature in a classroom
7. Year level
8. Distance from home to school (km)

::: {.callout-note .content-visible collapse="true" when-format="html"}
## Answers
1. Numerical, discrete (or ordinal — shoe sizes have a natural order but are
   not continuous)
2. Categorical, nominal
3. Numerical, continuous
4. Numerical, discrete
5. Categorical, nominal
6. Numerical, continuous
7. Categorical, ordinal (or numerical discrete depending on context)
8. Numerical, continuous
:::
