# Statistical Graphs and Displays {#sec-graphs}

A well-chosen graph makes patterns in data immediately visible. In this
chapter you will learn to **create**, **read**, and **critically interpret**
the most common statistical displays.

Before this chapter, make sure you are confident with @sec-data-types,
@sec-dependent-independent, and @sec-spread so your graph choices and
interpretations are statistically valid.

## Setting up our data

Throughout this chapter we use a simulated survey of 60 Year 11 students
across a New Zealand school. The data includes variables on transport,
height, study time, and maths scores.

```{r}
#| label: setup-data

set.seed(2024)
n <- 60

students <- data.frame(
  transport  = sample(c("Bus", "Car", "Walk", "Bike"), n,
                      replace = TRUE, prob = c(0.4, 0.25, 0.25, 0.1)),
  height_cm  = round(rnorm(n, mean = 167, sd = 9), 1),
  study_hrs  = round(pmax(0, rnorm(n, mean = 2.5, sd = 1.2)), 1),
  maths_score = round(pmin(100, pmax(0, rnorm(n, mean = 65, sd = 15))))
)
```

## Bar charts

Bar charts display the **frequency** (count) or **relative frequency**
(proportion) of categorical data.

```{r}
#| label: fig-bar-transport
#| fig-cap: "Bar chart showing how Year 11 students travel to school."

transport_counts <- table(students$transport)

barplot(
  sort(transport_counts, decreasing = TRUE),
  col  = c("#4dac26", "#b8e186", "#f1b6da", "#d7191c"),
  main = "How Year 11 Students Travel to School",
  xlab = "Mode of Transport",
  ylab = "Number of Students",
  border = "white",
  las = 1
)
```

### Reading a bar chart

::: {.callout-tip}
When describing a bar chart, always comment on:

1. **The most and least common** categories
2. **The overall pattern** (are categories roughly equal, or is one dominant?)
3. **The context** — what does this tell us about the group?
:::

**Sample answer:** "Bus is the most common way students travel to school,
with over 20 students using this method. Bike is the least common. More
students use public or active transport (bus, walk, bike) than passive
transport (car)."

## Dot plots

Dot plots are useful for **small numerical datasets**. Each data point is
shown as a dot above a number line.

```{r}
#| label: fig-dotplot-study
#| fig-cap: "Dot plot of daily study hours for Year 11 students."

# Rounded to nearest 0.5 for clearer dot plot
study_rounded <- round(students$study_hrs * 2) / 2

stripchart(
  study_rounded,
  method = "stack",
  pch    = 19,
  col    = "#0571b0",
  cex    = 1.4,
  xlab   = "Daily Study Hours",
  main   = "Daily Study Hours — Year 11 Students",
  offset = 0.4
)
```

## Histograms

A histogram shows the **distribution of continuous numerical data** by
grouping values into intervals (bins).

```{r}
#| label: fig-histogram-height
#| fig-cap: "Histogram of student heights (cm)."

hist(
  students$height_cm,
  breaks = 10,
  col    = "#2c7bb6",
  border = "white",
  main   = "Distribution of Student Heights",
  xlab   = "Height (cm)",
  ylab   = "Frequency",
  las    = 1
)
```

### Describing a histogram — SOCS

Use **SOCS** to describe a distribution:

| Letter | What to describe |
|--------|-----------------|
| **S**hape | Symmetric, skewed left, skewed right, uniform, bimodal |
| **O**utliers | Any unusual values far from the main group |
| **C**entre | Where the middle of the data is (mean or median) |
| **S**pread | How variable the data is (range, IQR, standard deviation) |

```{r}
#| label: fig-shape-examples
#| fig-cap: "Examples of different distribution shapes."
#| fig-height: 3

par(mfrow = c(1, 3))

# Symmetric
set.seed(1)
hist(rnorm(500, 50, 10), breaks = 15, col = "#4dac26", border = "white",
     main = "Symmetric", xlab = "", ylab = "Frequency", las = 1)

# Right-skewed
hist(rexp(500, 0.1), breaks = 15, col = "#f4a582", border = "white",
     main = "Right (positive) skew", xlab = "", ylab = "", las = 1)

# Left-skewed
hist(100 - rexp(500, 0.1), breaks = 15, col = "#92c5de", border = "white",
     main = "Left (negative) skew", xlab = "", ylab = "", las = 1)

par(mfrow = c(1, 1))
```

## Box plots (box-and-whisker plots)

A box plot summarises a **five-number summary**:
Minimum, Q1, Median, Q3, Maximum.

```{r}
#| label: fig-boxplot
#| fig-cap: "Box plot of maths scores by transport type."

boxplot(
  maths_score ~ transport,
  data   = students,
  col    = c("#4dac26", "#f4a582", "#92c5de", "#d7191c"),
  border = "#333333",
  main   = "Maths Scores by Transport Mode",
  xlab   = "Mode of Transport",
  ylab   = "Maths Score",
  las    = 1
)
```

### Parts of a box plot

```{r}
#| label: fig-boxplot-anatomy
#| fig-cap: "Anatomy of a box plot."
#| echo: false

set.seed(5)
x <- rnorm(100, 70, 12)
bp <- boxplot(x, horizontal = TRUE, col = "#2c7bb6", border = "#333333",
              main = "Parts of a Box Plot", xlab = "Value",
              axes = TRUE)

q <- quantile(x)
text(q[1], 1.3, "Minimum", cex = 0.85, col = "#d7191c")
text(q[2], 1.3, "Q1", cex = 0.85, col = "#d7191c")
text(q[3], 0.7, "Median", cex = 0.85, col = "#d7191c")
text(q[4], 1.3, "Q3", cex = 0.85, col = "#d7191c")
text(q[5], 1.3, "Maximum", cex = 0.85, col = "#d7191c")
```

## Stem-and-leaf plots

Stem-and-leaf plots show the actual data values and the shape of the
distribution simultaneously. They work well for **small datasets** (up to
~50 values).

```{r}
#| label: stem-leaf-scores

# Show a stem-and-leaf for a subset of maths scores
cat("Stem-and-leaf plot of maths scores (first 30 students):\n\n")
stem(students$maths_score[1:30])
```

## Scatter plots

Scatter plots show the **relationship between two numerical variables**.

```{r}
#| label: fig-scatter
#| fig-cap: "Scatter plot of study hours vs maths score."

plot(
  students$study_hrs, students$maths_score,
  pch  = 19,
  col  = adjustcolor("#0571b0", alpha.f = 0.7),
  xlab = "Daily Study Hours",
  ylab = "Maths Score",
  main = "Study Hours vs Maths Score — Year 11 Students"
)
abline(lm(maths_score ~ study_hrs, data = students),
       col = "#d7191c", lwd = 2)
```

### Describing a scatter plot

Comment on:

1. **Direction** — positive, negative, or no association
2. **Form** — linear or non-linear
3. **Strength** — strong, moderate, or weak
4. **Outliers** — any unusual points

::: {.callout-warning}
## Correlation ≠ Causation
Just because two variables are related **does not** mean one causes the other.
Always think about whether an association makes sense in context.
:::

## Pie charts

Pie charts show **proportions** of categorical data. Use them carefully —
bar charts are usually easier to read precisely.

```{r}
#| label: fig-pie
#| fig-cap: "Pie chart of transport modes."

pie(
  transport_counts,
  col    = c("#4dac26", "#f4a582", "#92c5de", "#d7191c"),
  main   = "Transport Modes — Year 11 Students",
  labels = paste0(names(transport_counts), "\n(",
                  round(100 * transport_counts / sum(transport_counts)), "%)")
)
```

## Choosing the right graph

| Situation | Best graph |
|-----------|-----------|
| One categorical variable | Bar chart, pie chart |
| One numerical variable (small n) | Dot plot, stem-and-leaf |
| One numerical variable (large n) | Histogram, box plot |
| Comparing groups numerically | Side-by-side box plots, back-to-back stem-and-leaf |
| Two numerical variables | Scatter plot |

## Check your understanding

1. A student recorded the heights of 25 classmates. Which graph type is most
   appropriate to display this data? Explain your choice.

2. Describe the scatter plot in @fig-scatter using the four features
   (direction, form, strength, outliers).

3. A bar chart shows the following transport counts: Bus = 24, Car = 15,
   Walk = 15, Bike = 6. What percentage of students walk to school?

4. True or false: "A right-skewed distribution has a longer tail on the left."
   Explain.

::: {.callout-note .content-visible collapse="true" when-format="html"}
## Answers
1. A histogram or dot plot — height is numerical and continuous.
2. Positive direction (more study → higher score), roughly linear form,
   moderate strength, possibly one or two outliers.
3. Walk = 15/60 = 25%.
4. False — a right-skewed (positively skewed) distribution has a longer tail
   on the RIGHT.
:::
