---
title: "Probability"
---

# Probability {#sec-probability}

**Probability** is a measure of how likely an event is to occur. It is
expressed as a number between 0 (impossible) and 1 (certain), or as a
percentage.

$$P(\text{event}) = \frac{\text{number of favourable outcomes}}{\text{total number of possible outcomes}}$$

## Probability scale

```{r}
#| label: fig-prob-scale
#| fig-cap: "The probability scale from impossible (0) to certain (1)."
#| echo: false
#| fig-height: 2

par(mar = c(3, 1, 2, 1))
plot(0, 0, xlim = c(-0.05, 1.05), ylim = c(-0.5, 0.5),
     type = "n", axes = FALSE, xlab = "", ylab = "",
     main = "Probability Scale")

# Draw scale
arrows(-0.02, 0, 1.02, 0, lwd = 2, code = 2, length = 0.1)
segments(c(0, 0.25, 0.5, 0.75, 1), -0.1, c(0, 0.25, 0.5, 0.75, 1), 0.1,
         lwd = 1.5)
text(c(0, 0.25, 0.5, 0.75, 1), -0.3,
     c("0\nImpossible", "0.25\nUnlikely", "0.5\nEven chance",
       "0.75\nLikely", "1\nCertain"),
     cex = 0.8)

# Colour gradient
for (i in seq(0, 1, by = 0.01)) {
  rect(i, 0.15, i + 0.01, 0.35,
       col = rgb(1 - i, 0, i, alpha = 0.6), border = NA)
}
```

## Key vocabulary

| Term | Definition |
|------|------------|
| **Experiment** | A process with uncertain outcomes |
| **Outcome** | A single possible result |
| **Event** | One or more outcomes grouped together |
| **Sample space** | The set of all possible outcomes |
| **Complementary event** | All outcomes that are NOT in the event |

## Simple probability — equally likely outcomes

When all outcomes are **equally likely**, we can use the classical definition.

### Example: rolling a die

```{r}
#| label: die-probability

sample_space <- 1:6
cat("Sample space:", sample_space, "\n")

# P(rolling a 4)
p_four <- sum(sample_space == 4) / length(sample_space)
cat("P(rolling a 4) =", p_four, "\n")

# P(rolling an even number)
p_even <- sum(sample_space %% 2 == 0) / length(sample_space)
cat("P(even number) =", p_even, "\n")

# P(rolling > 4)
p_gt4 <- sum(sample_space > 4) / length(sample_space)
cat("P(> 4) =", p_gt4, "\n")
```

## Complementary events

The **complement** of event A (written A' or $\bar{A}$) contains all
outcomes NOT in A.

$$P(A') = 1 - P(A)$$

**Example:** If the probability of rain tomorrow is 0.3, then:

$$P(\text{no rain}) = 1 - 0.3 = 0.7$$

```{r}
#| label: complement-example

p_rain <- 0.3
p_no_rain <- 1 - p_rain
cat("P(rain) =", p_rain, "\n")
cat("P(no rain) = 1 -", p_rain, "=", p_no_rain, "\n")
```

## Experimental (relative frequency) probability

When we cannot calculate probability theoretically, we **estimate** it by
repeating an experiment many times:

$$P(\text{event}) \approx \frac{\text{number of times event occurred}}{\text{total number of trials}}$$

### Simulating coin flips

```{r}
#| label: fig-coin-simulation
#| fig-cap: "Relative frequency of heads approaches 0.5 as the number of flips increases (the law of large numbers)."

set.seed(99)
flips  <- sample(c("Heads", "Tails"), 1000, replace = TRUE)
cumulative_heads <- cumsum(flips == "Heads") / seq_along(flips)

plot(
  cumulative_heads,
  type = "l",
  col  = "#0571b0",
  lwd  = 1.5,
  xlab = "Number of Flips",
  ylab = "Relative Frequency of Heads",
  main = "Simulated Coin Flips",
  ylim = c(0, 1),
  las  = 1
)
abline(h = 0.5, col = "#d7191c", lwd = 2, lty = 2)
legend("topright", c("Relative frequency", "True probability (0.5)"),
       col = c("#0571b0", "#d7191c"), lwd = 2, lty = c(1, 2), bty = "n")
```

::: {.callout-note}
## Law of Large Numbers
As the number of trials increases, the **relative frequency** (experimental
probability) gets closer and closer to the **true probability**. This is
why surveys and experiments need large samples.
:::

## Two-way tables

A **two-way table** (contingency table) organises data by two categorical
variables and is very useful for calculating probabilities.

```{r}
#| label: two-way-table

# Simulated data: 120 students, sport preference by gender
set.seed(7)
n_students <- 120
sport_data <- data.frame(
  Gender = sample(c("Male", "Female"), n_students, replace = TRUE,
                  prob = c(0.5, 0.5)),
  Sport  = sample(c("Rugby", "Netball", "Football", "Basketball"),
                  n_students, replace = TRUE, prob = c(0.3, 0.25, 0.25, 0.2))
)

tbl <- table(sport_data$Sport, sport_data$Gender)
tbl_with_total <- addmargins(tbl)
knitr::kable(tbl_with_total, caption = "Sport preference by gender (n = 120)")
```

### Reading probabilities from a two-way table

Using the table above:

```{r}
#| label: two-way-probs

total <- sum(tbl)

cat("P(Male) =", sum(tbl[, "Male"]), "/", total, "=",
    round(sum(tbl[, "Male"]) / total, 3), "\n")

cat("P(Rugby) =", sum(tbl["Rugby", ]), "/", total, "=",
    round(sum(tbl["Rugby", ]) / total, 3), "\n")

cat("P(Male AND Rugby) =", tbl["Rugby", "Male"], "/", total, "=",
    round(tbl["Rugby", "Male"] / total, 3), "\n")

cat("P(Rugby | Male) =", tbl["Rugby", "Male"], "/",
    sum(tbl[, "Male"]), "=",
    round(tbl["Rugby", "Male"] / sum(tbl[, "Male"]), 3), "\n")
```

The last probability — P(Rugby | Male) — is a **conditional probability**:
the probability of preferring rugby, **given** the student is male.

## Tree diagrams

Tree diagrams help calculate probabilities for **sequential events**.

```{r}
#| label: fig-tree-diagram
#| fig-cap: "Tree diagram for drawing two balls from a bag (with replacement)."
#| echo: false

# Simple tree diagram using base R
plot(0, 0, xlim = c(0, 4), ylim = c(0, 4), type = "n",
     axes = FALSE, xlab = "", ylab = "",
     main = "Tree Diagram: Drawing 2 Balls (Red/Blue, with replacement)")

# First branches
segments(0.5, 2, 1.5, 3.2, lwd = 1.5)  # Red
segments(0.5, 2, 1.5, 0.8, lwd = 1.5)  # Blue

# Second branches from Red
segments(1.5, 3.2, 2.8, 3.8, lwd = 1.5)
segments(1.5, 3.2, 2.8, 2.6, lwd = 1.5)

# Second branches from Blue
segments(1.5, 0.8, 2.8, 1.4, lwd = 1.5)
segments(1.5, 0.8, 2.8, 0.2, lwd = 1.5)

# Labels — first draw
text(0.9, 2.7, "P = 3/5", cex = 0.8, col = "#d7191c")
text(0.9, 1.3, "P = 2/5", cex = 0.8, col = "#0571b0")

# Node labels
text(1.5, 3.5, "Red", font = 2, col = "#d7191c")
text(1.5, 0.5, "Blue", font = 2, col = "#0571b0")

# Labels — second draw (from Red)
text(2.1, 3.7, "P=3/5", cex = 0.75, col = "#d7191c")
text(2.1, 2.8, "P=2/5", cex = 0.75, col = "#0571b0")

# Labels — second draw (from Blue)
text(2.1, 1.3, "P=3/5", cex = 0.75, col = "#d7191c")
text(2.1, 0.4, "P=2/5", cex = 0.75, col = "#0571b0")

# Outcome labels
text(3.6, 3.8, "RR: 9/25", cex = 0.8, font = 2)
text(3.6, 2.6, "RB: 6/25", cex = 0.8, font = 2)
text(3.6, 1.4, "BR: 6/25", cex = 0.8, font = 2)
text(3.6, 0.2, "BB: 4/25", cex = 0.8, font = 2)

# End nodes
text(2.8, 4.0, "Red", cex = 0.8, col = "#d7191c")
text(2.8, 2.4, "Blue", cex = 0.8, col = "#0571b0")
text(2.8, 1.6, "Red", cex = 0.8, col = "#d7191c")
text(2.8, 0.0, "Blue", cex = 0.8, col = "#0571b0")
```

**Key rule:** Multiply probabilities along each branch, add probabilities
for different paths leading to the same outcome.

## Expected value (extension)

For a simple scenario, the **expected value** is the long-run average outcome:

$$E(X) = \sum x \cdot P(x)$$

**Example:** A raffle has 200 tickets at $2 each. First prize is $100, second
prize is $50. What is the expected gain per ticket?

```{r}
#| label: expected-value

outcomes  <- c(100, 50, 0)         # winnings
probs     <- c(1/200, 1/200, 198/200)
cost      <- 2

expected_win  <- sum(outcomes * probs)
expected_gain <- expected_win - cost

cat("Expected winnings:", expected_win, "\n")
cat("Ticket cost:      ", cost, "\n")
cat("Expected net gain:", expected_gain, "\n")
cat("(A negative expected gain means this is not a good investment!)\n")
```

## Glossary

| Term | Definition |
|------|------------|
| **Probability** | Likelihood of an event, from 0 to 1 |
| **Sample space** | All possible outcomes |
| **Event** | One or more outcomes |
| **Complement** | All outcomes NOT in the event |
| **Relative frequency** | Experimental estimate of probability |
| **Conditional probability** | P(A given B): probability of A given B occurred |
| **Two-way table** | Table showing two categorical variables together |
| **Tree diagram** | Diagram showing sequential outcomes and probabilities |

## Check your understanding

1. A bag contains 4 red, 3 blue, and 3 yellow marbles. A marble is drawn at
   random. Find:
   a. P(red)
   b. P(not red)
   c. P(blue or yellow)

2. A class of 30 students was surveyed: 18 play sport, 12 play an instrument,
   and 6 do both.
   a. Draw a Venn diagram.
   b. Find P(plays sport but not instrument).
   c. Find P(neither sport nor instrument).

3. If P(A) = 0.4 and P(B) = 0.3 and A and B are independent, find P(A and B).

::: {.callout-note collapse="true"}
## Answers
1. a. 4/10 = 0.4; b. 6/10 = 0.6; c. 6/10 = 0.6
2. b. 12/30 = 0.4; c. (30 − 18 − 12 + 6)/30 = 6/30 = 0.2
3. P(A and B) = 0.4 × 0.3 = 0.12 (for independent events)
:::
