# Practice Questions and Achievement Standard Tips {#sec-practice}

This chapter brings everything together with practice questions at each
grade level, and advice on how to approach the AS91946 assessment.

## About the achievement standard

| Detail | Information |
|--------|-------------|
| **Standard number** | AS91946 |
| **Title** | Interpret and apply mathematical and statistical information in context |
| **Level** | NCEA Level 1 |
| **Credits** | 4 |
| **Assessment** | External (end-of-year exam) |
| **Strands** | Statistics and Probability |

## What each grade requires

### Achieved

- Correctly read values from graphs and tables
- Calculate basic statistics (mean, median, mode, range)
- State probabilities from given information
- Identify variable types
- Make simple statements about data in context

### Merit

- **Justify** your answers with reasons
- Compare and contrast groups using both centre AND spread
- Identify potential sources of bias
- Explain the difference between correlation and causation
- Interpret probabilities and relate them to real-world context

### Excellence

- Show **statistical insight** — go beyond the obvious
- Consider alternative explanations or limitations
- Make connections between different parts of the data
- Critically evaluate statistical claims
- Use appropriate statistical language precisely

## Writing tips for the exam

::: {.callout-tip}
## The golden rule
**Always refer back to the context.** Every answer should mention:
- The **variable** being described
- The **units** of measurement
- The **group** being described (population/sample)

❌ "The mean is 65."  
✅ "The mean maths score for Year 11 students was 65 out of 100."
:::

### Comparison sentence starters

| Purpose | Sentence starters |
|---------|------------------|
| Centre | "On average…"; "The typical…"; "The median…is higher/lower than…" |
| Spread | "The scores are more/less variable…"; "The IQR is larger, suggesting…" |
| Shape | "The distribution is roughly symmetric / right-skewed…" |
| Outliers | "There appears to be an outlier at…which may indicate…" |
| Causation | "However, this may be due to…rather than a direct causal relationship" |

## Practice set 1 — Achieved level

```{r}
#| label: practice-data-achieved
#| echo: false

set.seed(111)
heights <- round(rnorm(20, 168, 9), 1)
cat("Heights (cm) of 20 Year 11 students:\n")
cat(sort(heights), "\n")
```

**Questions:**

1. What type of data is height? Give a reason.
2. Calculate the mean height. Round to one decimal place.
3. Find the median height.
4. Find the range.
5. A student drew a histogram of this data. Circle the most appropriate
   description: **symmetric / right-skewed / left-skewed**

::: {.callout-note .content-visible collapse="true" when-format="html"}
## Worked solutions

1. Numerical, continuous — height can take any value within a range.

2. Mean: 164.8 cm

3. Median: 165.1 cm

4. Range: 37.3 cm

5. Likely symmetric (heights tend to be normally distributed).
:::

## Practice set 2 — Merit level

A local council surveyed 80 residents about their weekly exercise hours.
The results were:

```{r}
#| label: practice-data-merit

set.seed(222)
adults    <- round(pmax(0, rnorm(40, mean = 4.5, sd = 2.2)), 1)
teenagers <- round(pmax(0, rnorm(40, mean = 6.2, sd = 3.1)), 1)

summary_tbl <- data.frame(
  Group      = c("Adults", "Teenagers"),
  n          = 40,
  Mean       = round(c(mean(adults), mean(teenagers)), 2),
  Median     = c(median(adults), median(teenagers)),
  SD         = round(c(sd(adults), sd(teenagers)), 2),
  IQR        = round(c(IQR(adults), IQR(teenagers)), 2),
  Min        = c(min(adults), min(teenagers)),
  Max        = c(max(adults), max(teenagers))
)

knitr::kable(summary_tbl, caption = "Weekly exercise hours: adults vs teenagers")
```

```{r}
#| label: fig-merit-boxplot
#| fig-cap: "Weekly exercise hours for adults and teenagers."

exercise_df <- data.frame(
   group = rep(c("Adults", "Teenagers"), each = 40),
   hours = c(adults, teenagers)
)

boxplot(
   hours ~ group,
   data   = exercise_df,
   col    = c("#4dac26", "#0571b0"),
   border = "#333333",
   main   = "Weekly Exercise Hours",
   ylab   = "Hours per week",
   las    = 1
)
```

**Questions:**

1. Compare the centre of the two distributions. What does this tell us about
   exercise habits in context?
2. Compare the spread. What does this tell us?
3. The council used a telephone survey conducted at 2 pm on weekdays. Identify
   one source of bias and explain how it might affect the results.
4. Can we conclude from this data that teenagers exercise more because they
   are younger? Explain.

::: {.callout-note .content-visible collapse="true" when-format="html"}
## Model answers

**1. Centre:** The median exercise hours for teenagers (approx.
`r median(teenagers)` hrs) is higher than for adults (`r median(adults)` hrs).
This suggests teenagers in this sample tend to do more weekly exercise than
adults.

**2. Spread:** The IQR and standard deviation are larger for teenagers
(`r round(IQR(teenagers),1)` hrs vs `r round(IQR(adults),1)` hrs). This means
teenagers' exercise habits are more variable — some exercise a lot while others
exercise very little. Adults' exercise habits are more consistent.

**3. Bias:** Calling at 2 pm on weekdays means working adults are unlikely to
answer (they are at work). This leads to **non-response bias** — the adult
sample may over-represent retired or part-time workers, who may exercise more
than the general adult population.

**4. Causation:** No — we cannot conclude that age *causes* the difference in
exercise. Teenagers may exercise more because they participate in school sport,
not because of their age per se. Age and exercise are *correlated* in this
sample, but there may be confounding variables such as school PE requirements,
free time, or lifestyle factors.
:::

## Practice set 3 — Excellence level

A researcher claims: *"A study of 50 Auckland households found that households
with more bookshelves have higher household income. Therefore, buying more
bookshelves will increase your income."*

**Questions:**

1. Identify the two variables and their types.
2. Explain why this conclusion is flawed. Use statistical language.
3. Suggest one confounding variable that could explain the relationship.
4. What would the researcher need to do to properly establish causation?
5. The sample was 50 Auckland households selected by going door-to-door in
   Remuera (a wealthy suburb). Evaluate the reliability of generalising these
   findings to all New Zealand households.

::: {.callout-note .content-visible collapse="true" when-format="html"}
## Model answers

**1.** Number of bookshelves: numerical, discrete. Household income:
numerical, continuous.

**2.** The conclusion confuses **correlation with causation**. Even if the
two variables are positively correlated, this does not mean bookshelves
cause higher income. The relationship could be due to a third variable,
and correlation studies cannot establish cause-and-effect without
experimental control.

**3.** **Wealth/education level** — wealthier, more educated households are
more likely to own books/bookshelves AND to have higher income. Both
variables are driven by a common factor.

**4.** To establish causation, the researcher would need a **controlled
experiment** — randomly assigning households to receive bookshelves or not,
holding all other variables constant, and measuring income change over time.
This would be impractical.

**5.** The sample has significant **sampling bias**. Remuera is one of
Auckland's wealthiest suburbs. The sample does not represent the diversity
of New Zealand households in terms of income, region, or ethnicity. Results
cannot be reliably generalised to all New Zealand households. A
**stratified random sample** from across New Zealand would be needed.
:::

## Key formulas summary

```{r}
#| label: formulas-table
#| echo: false

formulas <- data.frame(
  Formula        = c("Mean", "Range", "IQR", "Probability", "Complement",
                     "Lower outlier fence", "Upper outlier fence"),
  Expression     = c(
    "x̄ = Σxᵢ / n",
    "Max − Min",
    "Q3 − Q1",
    "P(A) = favourable outcomes / total outcomes",
    "P(A') = 1 − P(A)",
    "Q1 − 1.5 × IQR",
    "Q3 + 1.5 × IQR"
  )
)

knitr::kable(formulas, col.names = c("Measure", "Formula"))
```

## Exam strategy checklist

Before the exam, make sure you can:

- [ ] Identify variable types (categorical/numerical, nominal/ordinal,
      discrete/continuous)
- [ ] Read values accurately from bar charts, histograms, box plots, and
      scatter plots
- [ ] Calculate mean, median, mode, range, IQR
- [ ] Describe distributions using SOCS (Shape, Outliers, Centre, Spread)
- [ ] Compare two groups using both centre and spread in context
- [ ] Calculate simple probabilities and use complementary events
- [ ] Read and calculate from two-way tables
- [ ] Identify bias in data collection methods
- [ ] Distinguish correlation from causation
- [ ] Critically evaluate statistical claims in media

## Useful resources

| Resource | Description | URL |
|----------|-------------|-----|
| NZQA | Official achievement standard information | [nzqa.govt.nz](https://www.nzqa.govt.nz) |
| CensusAtSchool | Real NZ student data for practice | [new.censusatschool.org.nz](https://new.censusatschool.org.nz) |
| iNZight | Free NZ-made statistics software | [inzight.nz](https://inzight.nz) |
| Stats NZ | Official NZ statistics and data | [stats.govt.nz](https://www.stats.govt.nz) |
| Khan Academy | Free video lessons on statistics | [khanacademy.org](https://www.khanacademy.org) |
