# Interpreting Statistical Information {#sec-inference}

A major focus of AS91946 is the ability to **critically read and interpret**
statistical information from a variety of sources — including newspapers,
websites, government reports, and scientific studies.

## Reading statistical claims

Statistical claims appear everywhere in everyday life. As a critical consumer
of information, you need to ask:

1. **Who** collected the data and why?
2. **How** was the data collected (survey, experiment, census)?
3. **How many** people or things were measured (sample size)?
4. **Who** was sampled — does the sample represent the population?
5. **What** do the numbers actually mean in context?

## Sources of bias

**Bias** is a systematic error that causes results to be unrepresentative.

| Type of bias | Description | Example |
|--------------|-------------|---------|
| **Sampling bias** | Sample does not represent population | Surveying only Year 11s about a school-wide issue |
| **Response bias** | People don't answer truthfully | "Do you cheat on tests?" |
| **Non-response bias** | Certain groups are less likely to respond | Online survey misses those without internet |
| **Leading question bias** | Question wording influences the answer | "Don't you agree that homework is too much?" |
| **Volunteer bias** | Only motivated people respond | Self-selected online polls |

```{r}
#| label: fig-bias-types
#| fig-cap: "Different types of bias and how they affect results."
#| echo: false

bias_types  <- c("Sampling", "Response", "Non-response", "Leading Q", "Volunteer")
bias_impact <- c(0.8, 0.6, 0.7, 0.9, 0.75)

barplot(
  bias_impact,
  names.arg = bias_types,
  col    = c("#d7191c", "#f4a582", "#fdae61", "#abdda4", "#2b83ba"),
  border = "white",
  main   = "Relative Impact of Bias Types (illustrative)",
  ylab   = "Severity (0 = low, 1 = high)",
  ylim   = c(0, 1),
  las    = 1
)
```

## Sample size and reliability

Larger samples generally give more **reliable** results — but size alone is
not enough if the sample is biased.

```{r}
#| label: fig-sample-size
#| fig-cap: "Variability in sample estimates decreases as sample size increases."

set.seed(42)
true_prop <- 0.55  # True proportion in population
sample_sizes <- c(10, 30, 50, 100, 200, 500, 1000)
n_simulations <- 1000

set.seed(321)
variability <- sapply(sample_sizes, function(n) {
  estimates <- replicate(n_simulations, mean(sample(c(TRUE, FALSE), n,
                                                    replace = TRUE,
                                                    prob = c(true_prop,
                                                             1 - true_prop))))
  sd(estimates)
})

plot(
  sample_sizes, variability,
  type = "b",
  pch  = 19,
  col  = "#0571b0",
  lwd  = 2,
  xlab = "Sample Size (n)",
  ylab = "Standard Deviation of Estimates",
  main = "Sample Size vs Variability of Estimates",
  las  = 1
)
abline(h = 0, col = "#d7191c", lty = 2)
```

## Correlation and causation

One of the most important concepts in statistical literacy is understanding
the difference between **correlation** and **causation**.

::: {.callout-warning}
## Correlation ≠ Causation
Two variables can be **correlated** (move together) without one causing
the other. There may be a **confounding variable** responsible for both.
:::

### Famous spurious correlations

- Ice cream sales and shark attacks both increase in summer → they are both
  caused by **hot weather**, not each other.
- Countries with more TVs have longer life expectancy → both are driven by
  **wealth**, not TVs causing people to live longer.

```{r}
#| label: fig-spurious
#| fig-cap: "A spurious correlation: ice cream sales and drownings both increase in summer."
#| echo: false

months <- 1:12
ice_cream <- c(30, 35, 50, 70, 90, 110, 120, 115, 85, 60, 40, 32)
drownings <- c(5, 6, 8, 12, 16, 20, 22, 21, 15, 10, 7, 5)

par(mar = c(5, 4, 4, 5))
plot(months, ice_cream, type = "b", pch = 19, col = "#d7191c", lwd = 2,
     xlab = "Month", ylab = "Ice Cream Sales (units)",
     main = "Ice Cream Sales and Drownings by Month",
     xaxt = "n", ylim = c(0, 140), las = 1)
axis(1, at = 1:12, labels = month.abb)

par(new = TRUE)
plot(months, drownings, type = "b", pch = 17, col = "#0571b0", lwd = 2,
     axes = FALSE, xlab = "", ylab = "", ylim = c(0, 25))
axis(4, las = 1, col.axis = "#0571b0")
mtext("Drownings", side = 4, line = 3, col = "#0571b0")

legend("topleft",
       c("Ice Cream Sales", "Drownings"),
       col = c("#d7191c", "#0571b0"), pch = c(19, 17), lwd = 2, bty = "n")
```

## Interpreting graphs critically

When reading statistical graphs, watch out for:

### Misleading y-axis

A y-axis that doesn't start at zero can make small differences look huge.

```{r}
#| label: fig-misleading-axis
#| fig-cap: "The same data displayed with a truncated y-axis (left) vs starting at zero (right)."

par(mfrow = c(1, 2))

years   <- 2020:2024
revenue <- c(980, 990, 995, 1005, 1010)

# Misleading
plot(years, revenue, type = "b", pch = 19, col = "#d7191c", lwd = 2,
     ylim = c(975, 1015), xlab = "Year", ylab = "Revenue ($M)",
     main = "Misleading: Truncated Axis", las = 1)

# Honest
plot(years, revenue, type = "b", pch = 19, col = "#4dac26", lwd = 2,
     ylim = c(0, 1100), xlab = "Year", ylab = "Revenue ($M)",
     main = "Honest: Starting at Zero", las = 1)

par(mfrow = c(1, 1))
```

### Percentage vs absolute numbers

::: {.callout-tip}
## Be careful with percentages
"Crime increased by 100%" sounds alarming — but if there were only 2 crimes
last year and 4 this year, the absolute increase is tiny.

Always ask: **100% of what?**
:::

### Averages without context

The **mean** can be pulled by extreme values, making it misleading if used
alone. Always ask what **measure of centre** was used and whether it is
appropriate.

## Reading media reports — a framework

Use these questions when evaluating a statistical claim in the media:

1. **What is being measured?** (the variable and its units)
2. **Who was studied?** (the population and sample)
3. **How was data collected?** (method and potential biases)
4. **How large was the sample?**
5. **What is the source?** (government, independent researcher, company with
   a vested interest?)
6. **Are conclusions supported by the evidence?** Or is the claim overstated?
7. **Are comparisons fair?** Are groups being compared actually comparable?

## A worked example

::: {.callout-note}
## News headline
*"Study shows eating chocolate improves exam scores by 15%"*
:::

**Critical analysis:**

- **What was measured?** Exam scores before and after eating chocolate.
- **Who was studied?** Not stated — red flag!
- **Sample size?** Not given — could be only 10 students.
- **Confounding variables?** Students who eat chocolate may be from wealthier
  families with better study resources.
- **Causation or correlation?** This is a correlational study; chocolate
  doesn't necessarily *cause* better scores.
- **Who funded it?** If a chocolate company funded it, there is a conflict of
  interest.

**Conclusion:** The headline is likely exaggerated or misleading. We would
need much more information before accepting this claim.

## Glossary

| Term | Definition |
|------|------------|
| **Bias** | Systematic error that distorts results |
| **Confounding variable** | A hidden variable that explains an apparent relationship |
| **Correlation** | A statistical relationship between two variables |
| **Causation** | One variable directly causes a change in another |
| **Reliable** | Consistent results if repeated |
| **Valid** | Measuring what it claims to measure |
| **Population** | The full group of interest |
| **Sample** | A subset of the population |

## Check your understanding

1. A newspaper article claims: "People who eat breakfast earn more money than
   those who skip it." Identify a possible confounding variable.

2. A survey found that 80% of students support a longer lunch break. The
   survey was conducted by standing outside the school canteen. What type of
   bias is this? How might it affect results?

3. A graph shows the number of electric vehicles sold in New Zealand, with the
   y-axis starting at 45,000 rather than 0. Sales increased from 47,000 to
   51,000. On the graph, this looks like a 400% increase. Calculate the true
   percentage increase.

4. Explain the difference between correlation and causation using your own
   example.

::: {.callout-note .content-visible collapse="true" when-format="html"}
## Answers
1. Wealthier people are more likely to eat breakfast (they can afford food)
   and also earn more — wealth is the confounding variable.
2. Sampling bias / volunteer bias — students near the canteen are more likely
   to want a longer lunch break. Results will overestimate support.
3. True % increase = (51,000 − 47,000) / 47,000 × 100 ≈ 8.5%.
4. Accept any reasonable example, e.g., "Countries with more hospitals have
   higher death rates — hospitals don't cause death; both are linked to
   population size."
:::
